Post-replication citation patterns in psychology: Four case studies

### 2021-07-19 manuscript/manuscript.Rmd
Update manuscript.Rmd

output            : 
  papaja::apa6_word
---

@@

### Citations to replication studies and co-citation of original studies

@@

## Article characteristics

@@

Discussion
It has been proposed that replication studies can facilitate scientific self-correction by modifying scientists’ belief in the credibility of published findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018); however, the extent to which this occurs in practice is unclear. In this study, we investigated how the research community responded to four strongly contradictory replications in the field of psychology by examining post-replication citation patterns for original studies. We observed some ‘progressive’ response patterns in the form of modest active correction (a small decline in favourable citations and a small increase in unfavourable citations), but also even more prominent ‘regressive’ response patterns in the form of unchallenged belief perpetuation (sustained levels of favourable citations and few unfavourable citations, particularly in the Baumeister case) and considerable citation bias (neglecting to cite the replication study; in all cases aside from Sripada). When authors cited the original study favourably despite the replication result, only half of the articles provided any explicit counter arguments in defence of the original study (an explicit defence). Overall, these findings are consistent with prior observations that favourable citation patterns appear relatively unperturbed by subsequent publication of contradictory results in studies with lower risk of bias (Tatsioni et al., 2007) or even by full retraction (Budd et al., 1998; Fernandez & Vadillo, 2020), and that positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence once a theory gets entrenched, despite overwhelming evidence against it (Bastiaansen et al., 2015; Greenberg, 2009).

@@

```{r child = "appendix.Rmd"}
```

>>>>

output            : 
  papaja::apa6_pdf:
    includes:
        after_body: "appendix.tex"
---

@@

@@

@@

## Citations to replication studies and co-citation of original studies

# Discussion
It has been proposed that replication studies can facilitate scientific self-correction by modifying scientists’ belief in the credibility of published findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018); however, the extent to which this occurs in practice is unclear. In this study, we investigated how the research community responded to four strongly contradictory replications in the field of psychology by examining post-replication citation patterns for original studies. We observed some ‘progressive’ response patterns in the form of modest active correction (a small decline in favourable citations and a small increase in unfavourable citations), but also even more prominent ‘regressive’ response patterns in the form of unchallenged belief perpetuation (sustained levels of favourable citations and few unfavourable citations, particularly in the Baumeister case) and considerable citation bias (neglecting to cite the replication study; in all cases aside from Sripada). When authors cited the original study favourably despite the replication result, only half of the articles provided any explicit counter arguments in defence of the original study (an explicit defence). Overall, these findings are consistent with prior observations that favourable citation patterns appear relatively unperturbed by subsequent publication of contradictory results in studies with lower risk of bias (Tatsioni et al., 2007) or even by full retraction (Budd et al., 1998; Fernandez & Vadillo, 2020), and that positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence once a theory gets entrenched, despite overwhelming evidence against it (Bastiaansen et al., 2015; Greenberg, 2009).

@@

```{r render_appendix, include=FALSE}
render_appendix("appendix.Rmd")
```
